### 2014-02-27 paper/abstract.tex
more paper text, algo and proof for neon

\begin{abstract}
3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.  However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.  While macro-scale digital fabrication devices cannot manufacture electronics in-place, there have been many clever approaches to redirecting interaction onto 3D prints.  We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.  We present a technique and design tool for routing tubes of various topologies through the interior of 3D printed parts.  Our design tool is integrated into a 3D model manipulation program.  It allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.  We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.
\end{abstract}

>>>>

\begin{abstract}
3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.  However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.  While macro-scale digital fabrication devices cannot manufacture electronics in-place, we propose a general technique for redirecting active components through the core and onto the surface of 3D printed interactive objects using a series of tubes.  We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.  We present a technique and design tool for routing tubes of various topologies through the interior of 3D printed parts.  Our design tool is integrated into a 3D model manipulation program.  There are two distinct routing algorithms.  One allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.  The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of its optimality.  We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.
\end{abstract}

### 2014-02-27 paper/designspace.tex
more paper text, algo and proof for neon

@@

\valkyrie{we can just include related work in this section?  we basically cover it in the table, anyway.}

\begin{figure*}[t]
\centering
    \includegraphics[width=\textwidth]{figures/designspace.png}
\caption{The design space of tube-based interactions.  Existing systems are written in regular font.  Those created with fabrication are in {\color{blue}blue}. In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.  Darker grey is output.  Lighter grey is input.  \valkyrie{I don't know what else might go in the blank spaces... looking for suggestions!  Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...}}
\label{fig-designspace}
\end{figure*}

\valkyrie{optimally, we can just describe this thoroughly in the chart, and avoid spending a bazillion paragraphs talking about each thing individually here}

\subsection{Inputs}

\subsubsection{Flexing}

Much like \cite{Slyper-shape}, we can sense flexing and bending of prints made on the Objet.  We can make prints on the Objet and tunnel through them, though, without making crazytastic silicone molds.  We can make things flex using muscle wire and cleverly-placed expandable air pockets.

\subsubsection{Touch}
Capacitance (digital).  SFCS.

\subsubsection{Pressure}
Pressure (via SFCS or fluid pressure).

\subsubsection{Tapping}
 Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.  We use hard tubes in soft things, anyway, for the conductive goo.

\subsubsection{Other stuff?}
Could be.

\subsubsection{Traditional Components}
Obviously you can hook up traditional buttons, etc., the same way as always.  With copper paint instead of traditional wiring, we can share grounds, and we don't have to solder.

\subsection{Outputs}

\subsubsection{Visual}
EL wire.  Colored liquids.  Mechanical motion by pushing light or tube-plugging (like a big ball) objects with fluid.

\subsubsection{Aural}
Resonance chambers.  Air cavity design for sound/amplification (see passive iPhone speakers).  I mean, this is basically just 3D printing instruments, which we know has been done.

\subsubsection{Haptic}
Compressible and incompressible fluids for actuation.  Recreation of PneUIs.  Tactile output.  Adding particles to add extra feedback.

\subsubsection{Olfactory/Gustatory}
Different chambers filled with different scented/flavored fluids.  We can mix them using pipes and pressure.

@@

Both identification by recall and intentional encoding are possible.  As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.  The presence of tubes changes this, thus allowing for recall of a shape once its acoustic signature has been recorded.  In addition, semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).  All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the tube.

@@

>>>>

We propose the use of tubes to redirect active components through the core and onto the surface of 3D printed interactive objects.  By using different types and topologies of tubes and inserting different media into the completed tubes, a variety of input and output types \valkyrie{need clearer word than ``types"} can be located at arbitrary points on the surface of a 3D print.  We describe the design space of tubes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.

@@

\subsection{Interaction Techniques}
These didn't get discussed yet.
\begin{itemize}
\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
\end{itemize}

\valkyrie{this section includes related work.  is that ok?  it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).}

By crossing tube types, topologies, and inserted media, we create a space of possible inputs and outputs that can be tube-mediated.  We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication.  We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.

\begin{figure*}[t]
\centering
    \includegraphics[width=\textwidth]{figures/designspace.png}
\caption{The design space of tube-based interactions.  Existing systems are written in regular font.  Those created with fabrication are in {\color{blue}blue}. In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.  Darker grey is output.  Lighter grey is input.  \valkyrie{I don't know what else might go in the blank spaces... looking for suggestions!  Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...  and the general presentation of the table could probably be better in some way.}}
\label{fig:designspace}
\end{figure*}

\subsubsection{Inputs}

Tubes create opportunities for many types of input sensing across the surface of printed devices.  We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.  We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of tubes.

Touch sensing is enabled through tubes filled with conductive media.  Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).  Using conductive paint and Swept Frequency Capacitve Sensing (SFCS \cite{Sato-touche}), an interior star tube topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).  Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.

Pressure can be sensed in multiple ways.  Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed fabricatable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.  \valkyrie{that sentence is not clear}  These primitives can be attached at the terminus of any semi-closed tube; the user can manipulate the printed endpoint and the system can sense air pressure changes.  Pressure can also be more simply sensed via capacitance using those techniques.

Grasp sensing can also be enabled using Wimmer's FlyEye technique \cite{Wimmer-flyeye}.  This technique involves a pair of optical links at each point of interest: one of each pair is connected to an infrared (IR) LED and the other to a camera.  When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.  Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely dense particulates enclosed in a flexible volume, and can be sensed optically.  These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.  The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow tubes (see Figure \ref{fig:pens}).

Some 3D printers, like our Objet Connex 260, can fabricate rubber-like materials.  Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.  While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing tubes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.

Tapping is another input possibility.  Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), tubes of a greater conducting material can be embedded in a model of a lesser conducting material.  A microphone or piezo placed at the system-side of these sound-conducting tubes can determine where the model was tapped and how it was tapped.  Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting tubes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.

Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.  Components can be recessed into a print's surface, with their leads implanted into open tubes.  By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:embeddedcomponents}).

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{This sphere has a recessed space to hold a potentiometer.  The potentiometer and the LEDs share a ground, seen by the translucent rendered view in (b).}
\label{fig:embeddedcomponents}
\end{figure}

\subsubsection{Outputs}

While many input techniques require the use of conductive media inserted into tubes, output media are more varied.  We describe potential tube-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.  In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.

\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.  EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.  Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for dispaly.  PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a tube could be actuated by a system-side air pump (see Figure \ref{fig:braille}).  Mechanical motion can be mediated through fluids (a light object which is partially contained in a tube or located at its end could be pushed by the fluid pressure of the tube) or threadables (e.g., muscle wire).  Large particles trapped in tubes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).  Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.  I think all these paragraphs might seem terrible...}

\emph{Aural} outputs can be created with gases.  Zoran has successfully 3D printed a functional flute, which in essence is a long open open tube with semi-closed tubes coverable by keys \cite{Zoran-flute}.  Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.  All that is necessary for this is an enclosed chamber with a tube connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).  Passive sound amplification (like a phonograph horn) and sound redirection are also possible using tubes.

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).  This resonator creates a sound when air is blown through the tube at the top, because air compresses at the small neck joint between the chamber and the tube.  \valkyrie{need to get a test print of this that actually works}}
\label{fig:embeddedcomponents}
\end{figure}

\emph{Haptic} outputs are possible through compressible and incompressible fluids.  Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).  Additionally, use of semi-closed tubes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:braille}).  Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of tubes.

\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of tubes carrying scented or flavored fluids, respectively.

@@

Both identification by recall and intentional encoding are possible.  As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.  Additionally, two objects that are visually identical but which have fully enclosed (or other types) of tubes on their interior can be distinguised acoustically.  Thus we can recall an object's identity once its acoustic signature has been recorded.

We have also experimented with intentional encoding.  Semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).  All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the tube. \valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.  my sense is that this will depend highly upon the i/o sensitivities of the mic and speaker selected.}

@@

This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.  While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.

### 2014-02-27 paper/relatedwork.tex
more paper text, algo and proof for neon

Our work is related to three existing research areas: fabrication, interaction techniques, and routing.

@@

Previous work has investigated the integration of interactivity into the fabrication process.  \cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays.  \cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
\begin{itemize}
\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
\item \cite{Willis-infrastructs} - InfraStructs, identification of 3d printed objects using Terahertz imaging
\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.  input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.  they don't offer a routing algorithm or anything, though.
\end{itemize}

\subsubsection{Printing Electronics}
\begin{itemize}
\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
\item \cite{Optomec} - fancy machines that spray conductive film
\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
\end{itemize}

\subsection{Routing}
\begin{itemize}
\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
\item \cite{Park-microchannels} - similar to injection of liquid metal, above
\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
\end{itemize}

\subsection{Interaction Techniques}
\begin{itemize}
\item \cite{Sato-touche} - Touch\'{e}.  It's like Touch\'{e} with sound, but without sound.  SFCS.
\item \cite{Harrison-buttons} - fabricate latex + acrylic buttons and pressurize with air
\item \cite{Sodhi-aireal} - air vortex generation in free space.  air haptics.
\item \cite{Yao-pneui} - PneUIs, creating interfaces with pneumatics
\item \cite{Slyper-shape} - creating silicone bendy things with embedded electronics to sense flexing, stretching, etc., supported by those shapes.
\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
\item \cite{Ono-touchandactivate} - the Touch\'{e} with sound paper from last year's UIST
\item \cite{Follmer-jamming} - Jamming User Interfaces
\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
\end{itemize}

>>>>

While using a series of tubes can replicate the interaction techniques discussed (and more), our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.

@@

Previous work has investigated the integration of interactivity into the fabrication process.  \cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays.  \cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
\begin{itemize}
\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.  input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.  they don't offer a routing algorithm or anything, though.
\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
\item \cite{Optomec} - fancy machines that spray conductive film
\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
\end{itemize}

\subsection{Routing}
\begin{itemize}
\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
\item \cite{Park-microchannels} - similar to injection of liquid metal, above
\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance

Our neon routing algorithm is similar to that described by Wong, et al., in \cite{Wong-continuousline}.  However, We are not confined to the plane to avoid line-crossings, and we also have additional freedom in creating lines where none existed.  Because we can shield neon post-print with black tape or material (thus rendering it invisible), we don't have the same need to avoid drawing new line segments, and we are free to create much shorter paths.  It's not quite a TSP
\end{itemize}

### 2014-02-27 paper/tubes.tex
more paper text, algo and proof for neon

\subsection{Tool for Tube Design}

@@

\subsubsection{Selection}

\textbf{Exterior Connection Points}

@@

\textbf{Designing Interior Paths}

@@

\subsubsection{Mesh Modification}

>>>>

\section{Tool for Tube Design}

@@

\subsection{Exterior Connection Points}

@@

@@

\subsection{Designing Interior Paths}

Our problem here is a relaxed version of the Chinese Postman Problem$^{[\valkyrie{citation needed}]}$ in which we wish to traverse all edges of the graph described in the user's input (as the edges of the graph are the lines the user wishes to have lit) with a single circuit.  It is a relaxation in that we allow the creation of non-existing paths: if components are disconnected (e.g., a user wishes to spell ``UIST" in neon lights), we must create edges that connect them; additionally we can create edges to connect odd-degree vertices rather than simply retracing existing edges.  In the final design, all created edges will be shielded by dark material so the inserted medium is not visible.

We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.

Let $G=(V,E_0)$ s.t. $\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.  Let $e_{temp} = (start, end)$, $E' = e_{temp} \cup E$.

We need to connect disconnected subgraphs in $G$.  Let $G_{dis} = \{G_1, G_2, ... G_n\} \in G$ s.t. $\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.  Create $E_{conn} = \{e = (u, v), u \in G_i, v \in G_j, i \neq j, weight(e) = distance(u, v)\}$ be a set of edges.  Sort $E_{conn}$ s.t. $E_{conn} = \{e_1, e_2, ... e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.  Let $E_{conn-min} = \{ \}$.

Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.  Then let $E_{dupe} = \{ e = (u, v) \in E_{conn}$ s.t. $u$  is reachable from $v$ along edges $E' \cup E_{conn-min}$.  Let $E_{conn} = E_{conn} \setminus E_{dupe}$.  Repeat until $E_{conn} = \{ \}$.  Let $E' = E' \cup E_{conn-min}$.

At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.  Let $V_{odd} = \{ v \in V s.t. degree(v) \% 2 = 1 \}$.  Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t. $\forall e=(u,v) \in E_{circuit}, weight(e) = distance(u, v)$.  Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.  Let $E' = (E\setminus e_{temp}) \cup E_{circuit-min}$.

\begin{theorem}
 $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
\end{theorem}
\begin{proof}
$E \subseteq E'$ : we never remove edges from $E$ in our algorithm.  $\therefore E \subseteq E'$.

$G' = (V, E')$ is connected : if $G'$ is not connected, $\exists G_i = (V_i, E_i) \subset G', u \in V_i, v\in V \setminus V_i$ s.t. $u$ is not reachable from $v$.  $G_i \not\in G_{dis}$, because we create edges connecting $G_j, G_k \forall G_j, G_k \in G_{dis}$ and only remove an edge $e_{dupe} = (u, v)$ from $E_{conn}$ once we determine that $u$ is reachable from $v$ along edges $E' \cup E_{conn-min}$.  $\therefore G_i \not\in G_{dis}$.  This implies that $G_i$ was not initially disconnected from $G$, because by definition $G_{dis}$ is the set of all disconnected subgraphs of $G$.  We cannot have disconnected $G_i$ from $G$ because $E \subseteq E'$.  Thus, a contradiction.  $\therefore G' = (V, E')$ is connected.

$G' = (V, E')$ is semi-Eulerian : Each edge $e$ in our minimum weight matching connects a unique pair of vertices $v_i, v_j \in V_{odd}$ by the definition of minimum weight matching.  Each edge $e = (v_i, v_j) \in E_{circuit-min}$ adds one to the degree of $v_i$ and $v_j$, causing them to be of even degree. $|V_{odd}| \% 2 = 0$ by the handshake lemma, $\therefore$ all edges can be paired.  When we remove $e_{temp} = (start, end)$ from $E'$, we cause those vertices, which were of even degree by the above process, to be of odd degree.  Thus, all vertices except $start$ and $end$ (which are of odd degree) are of even degree.  $\therefore G'$ is semi-Eulerian.
\end{proof}

\subsection{Mesh Modification}
